Machine Learning Manuscript Outline

### 2019-01-17 submission/ML_outline.Rmd
for Pat and Jenna

---
title: "Machine Learning Manuscript Outline"
output: word_document
---

@@

##### General Context of the work
As the microbiome field continues to grow, there is an ever-increasing demand for reproducible methods for identifying associations between members of the microbiome and phenotypes. Currently, the field’s use of machine learning lacks clarity and consistency.

@@

* One application of machine learning to microbiome data has been to classify patients as having colorectal tumors based on microbiota-associated biomarkers. Colorectal cancer is one of the leading cause of death among cancers in the United States. Early diagnosis increases the chance of survival. However the current diagnostic methods are expensive and invasive. As a less invasive tool,  studies use relative abundances of the gut bacteria populations to predict disease progression.

* Most microbial communities are pretty patchy and the likelihood of a single feature that explains the differences in health is pretty small. It is likely that many biomarkers are needed to account for the patchiness as well as the context dependency of the features.

##### Identification of a gap or other need for research
* ML use in microbiome literature is a bit like the wild west with lack of clarity over methods, testing, validation, etc. There is a need for guidance on how to properly implement these different methods.
* The prediction performance of previous CRC microbiome models vary greatly, with areas under the receiver operating characteristic curve (AUC) of 0.7-0.9.

##### Summary of appraoch to answer the research question
* We establish a non-leaky pipeline. 
* We perform L2-regularized logistic regression, Linear L1 and L2 SVM, Non-Linear SVM, Decision tree, Random forest, XGBoost. 
* We evaluate the classification success of different machine learning methods and discuss the reproducibility, robustness, actionability, interpretibility and susceptibility to overfitting of each method.

##### Announcement of principal findings
* Prediction perfomance of each model. 
* Is there a maximum threshold of prediction with all these methods?
* Does an increase in model complexity improve predictibility?
* Synthesis statement regarding modeling 16S microbiome data

##### Analysis of data
* What are the features and what are the labels?
We used the fecal hemoglobin concentration and 16S rRNA gene sequences from stool samples to classify 490 patients as having advanced tumors (advanced adenoma or carinoma) or not (non-advanced adenoma or normal colon). 
* What is the data (temporal or not)? Assumptions we make when we use a dataset. What will the future data look like?
* Pre-proccessing of the data
* Machine Learning pipeline backbone. How do I split, train, validate and test?
A diagram to explain the modeling pipeline.
* Which methods are linear/non-linear? Talk about interpretibility.
* Cross-validation and hyper-parameter tuning methods for each modeling method
* Programming languages and packages/modules utilized
* Statistical methods for comparison of model performance

@@

##### AUC results of 7 modeling approaches. (FIT + OTUs)

@@

##### Hyper-parameter tuning budgets and corresponding AUC values.

@@

@@

* The interactions between the biomarkers may be nonlinear. Obviously, the linear models will not incorporate this because they are linear. Tools like linear models (e.g. metastats, lefse, wilcoxon, etc) are likely worthless.

@@

@@

>>>>

---
title: "Machine Learning Manuscript Outline"
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
    includes:
      in_header: header.tex
csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
fontsize: 11pt
geometry: margin=1.0in
---

@@

##### General Context of the work

* As the microbiome field continues to grow, there is an ever-increasing demand for reproducible methods for identifying associations between members of the microbiome and human health.

* Most microbial communities are pretty patchy and the likelihood of a single species that explains the differences in health is pretty small. It is likely that subsets of those communities in relation to one another are needed to account for the patchiness as well as the context dependency of the features.

* Thus, researchers have started to explore the utility of machine learning (ML) techniques.

* Currently, the field’s use of machine learning lacks clarity and consistency on
    1. Which methods are used?
    2. Why those methods are used? Can these methods explain if there are linear or non-linear relationships between phenotype and features?
    3. How those methods are used?
    4. Reproducibilty

* ML methods have a trade-off for interpretibility, susceptibility to overfitting and actionability versus prediction performance. It is important to prioritize and define task accordingly.

@@

* Among cancers, colorectal cancer is one of the leading causes of death in the US. While colonoscopy is an effective screening tool, it is invasive and as a result has a low rate of patient adherence. In contrast, microbiome analysis of stool is non-invasive. Machine learning (ML) techniques are used to identify patients with colorectal tumors based on microbiota-associated biomarkers.

* However, the discriminative performance of these models varies greatly, with areas under the receiver operating characteristic curve (AUROC) of 0.7-0.9.

* Previously:
    - Differences in task definition (what is it that we want to predict?)
    - Use of logistic regression without discussion of how it explains non-linearity of microbiome & CRC.
    - Use of random forest only without discussion of interpretibility.
    - Use of random forest without proper pipeline.

* To shed light on how much differences in modeling can affect the results, we performed an empirical analysis comparing several different modeling pipelines.

##### Summary of appraoch and findings

* Modeling pipelines were established for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost.

* These methods increase in complexity while decrease in interpretibility.

* We established ML pipeline with held-out test data and performed 100 data-splits to evaluate generalization and predicton performance of the ML method.

* Applied to held-out test data, the mean AUROC varied from 0.68 (std ± 0.04) to 0.82 (std ± 0.04).

* Random Forest had the highest mean AUROC for detecting SRN and was less susceptible to overfitting compared to other methods.

* Despite the lower mean AUROC value, the L1-regularized linear kernel SVM offered the greatest interpretability and stability.

* In terms of computational efficiency, x trained the fastest, while y took the longest.

* We found that cross-validation and testing AUROC could vary by as much as 0.06, highlighting the importance of a separate held-out test set for evaluation.

* Aside from evaluating generalization and classification performance for each of these models, this study established standards for modeling pipelines of microbiome-associated machine learning models.

@@

##### AUROC results of 7 modeling approaches. (FIT + OTUs)

@@

##### Hyper-parameter tuning budgets and corresponding AUROC values.

@@

* Emphasize trade-off results we observe in this study.

@@

* The interactions between the biomarkers may be nonlinear. Obviously, the linear models will not incorporate this because they are linear. Tools like linear models (e.g. metastats, lefse, wilcoxon, etc) are likely worthless. How many OTUs would you find with these methods?

@@

* Why did we use only one dataset? Becuase we are comparing several methods on 1 dataset.

@@

* Baxter et al, 2016

##### Analysis of data

* What are the features and what are the labels?

    1.  Features: Fecal hemoglobin concentration and 16S rRNA gene sequences from stool samples

    2.  Labels:  490 patients as having advanced tumors (advanced adenoma or carinoma) or not (non-advanced adenoma or normal colon).

* What is the data (temporal or not)? Assumptions we make when we use a dataset. What will the future data look like?

* Pre-proccessing of the data

* Machine Learning pipeline backbone. How do I split, train, validate and test?

      - A diagram to explain the modeling pipeline.

* Which methods are linear/non-linear? Talk about interpretibility.

* Cross-validation and hyper-parameter tuning methods for each modeling method

* Programming languages and packages/modules utilized

* Statistical methods for comparison of model performance

* Code/data availability
